From noise to knowledge: how randomness generates novel phenomena and information

### 2017-10-10 paper.Rmd
WIP

- Demographic stochasticity 
- Environmental stochasticity
- Individual heterogeneity
- Spatial variability
- stochastic chaos

@@

## Noise the creator: Noise can induced novel phenonema

- Oscillations (parametric oscillator, stochastic resonance)
- population inflation
- Critical transitions
- Spatial pattern formation
- Population persistence (bet-hedging)
- Species coexistence

@@

>>>>

**Demographic stochasticity**.  Demographic stochasticity refers to fluctuations in population sizes or densities that arise from the fundamentally discrete nature of individual birth and death events.  Demographic stochasticity is a particularly instructive case for illustrating a mechanism for how noise arises as an aggregate description from a fundamentally lower-level mechanistic process.  Demographic stochasticity is typically modeled as Markov jump process: in a population of $N$ identical individuals at time $t$, a birth occurs with probability $b_t(N_t)$ (_i.e._ a rate that can depend on both the population size $N$) and increases the population size, $N$ to $N+1$, and deaths occur with probability $d_t(N_t)$, decreasing the population size, $N$ to $N-1$.  Assuming each of these events are independent, this is a Poisson process

[@Black2012]

**Environmental stochasticity**

[@Black2012]

**Individual heterogeneity**

**Spatial variability**

### Not noise

In addition to these sources of variation, there are additional sources of variation or uncertainty in a model which we will not consider for the purpose of this review which, while related, operate in a fundamentally different way that does not give rise to the same phenomena and information.

**Observational error**: Observational error acts much like another form of stochasticity, arising from
sampling effects or limitations of instruments, etc. Yet unlike the sources described above, this 
uncertainty does not impact the actual state of our system, and thus does not give rise to any
additional phenomena or information about that process.

**Reducible uncertainty**: In constructing and estimating mathematical models of ecological processes, 
we encounter uncertainty over our parameter estimates and possible choice of models [@Ellison2004].  In some ways, this
uncertainty can act like the stochasticity described above: for instance, both introduce variation into
any forecast of future outcomes. Yet unlike stochasticity, this uncertainty is reducible: with additional
observations, we can narrow the uncertainty over parameters or models, while stochasticity remains an
inherent source of variation.

**Chaos**: Deterministic chaos [@Desharnais2001] is another mechanism of introducing significant variation into a simple
model which can be observationally indistinguishable from a stochastic process [@Werndl2009].

Figure: schematic showing types of noise

@@

Figure:

## Noise the creator: Noise can induce novel phenonema

- Oscillations (parametric oscillator, stochastic resonance)

[@Black2010]

- population inflation

@Read2007 <!-- Stochasticity generates an evolutionary instability for infectious disease -->

- Critical transitions

[@Schreiber2008] decreasing variance before crash

- Spatial pattern formation
- Population persistence (bet-hedging)
- Species coexistence

Figure:

@@

Figure:

## Conclusions
